A Bayesian Framework for Learning Regular Expressions

### 2012-12-09 paper/paper.tex
Added small corrections.

\subsection{Likelihood}
To define the likelihood of producing certain strings, we turn a DFA into a generative automaton with a probability distribution over generated strings. Letting the input strings be independent, the probability of generating the set of strings $S$ is equal to the product of the probabilities of generating each string.
\begin{align*}
	p(S) = \prod_{s \in S} p(s)
\end{align*}

Let the string $s$ be the concatenation of characters $\{a_0,a_1,...,a_{n-1}\}$. We define the probability of generating a string as the product of the probability of generating the letters in sequence $p(a_0,a_1,...,a_{n-1})$ and the product of stopping the string generation on an accept state $p(\text{stop})$.
\begin{align*}
	p(s) = p(a_0,a_1,...,a_{n-1})p(\text{stop})
\end{align*}

We define the probability of generating the characters as the product of the probabilities of making each state transition.
\begin{align*}
	p(a_0, a_1, ..., a_{n-1}) = \prod_{i=0}^{n-1}p(\delta(r_i, a_i))
\end{align*}
where $r_i$ is the state reached after the first $i$ characters, and $r_0$ is the start state.

To simplify inference, we assume uniform probability over all possible outbound transitions for a given state. Formally, the probability of making a state transition is given by:
\begin{align*}
	p(\delta(r_i, a_i)) &= \frac{1}{\sum_{a\in A} 1\{\delta(r_i, a)\in Q\}}
\end{align*}
where $1\{\cdot\}$ is the indicator function, which evaluates to $1$ if the enclosed proposition is true and $0$ otherwise. $\delta(r_i, a)\in Q$ indicates that the state transition leads to another state of the DFA, so the denominator counts the number of characters for which a transition is defined. In the case of the wildcard `A', the transition is defined for all 26 letters, so the denominator is 26.

@@

Instead of leaving $\tau$ as a parameter of the model, we assume a uniform prior over the parameter and marginalize it when calculating the likelihood of a string. So, let $b$ be the number of times the DFA skipped an accept state before accepting. That is, $b$ is given by the equation below, where $F$ is the set of accept states.
\begin{align*}
	b = \sum_{i=0}^{n-1} 1 \{ r_i \in F\}
\end{align*}

% TODO: Cite this

The probability of stopping at an accept state is given by the equation below. The condition $r_n \in F \wedge \exists a : \delta(r_n, a)\in Q$ states that the last state reached by the string is an accept state, and that that state has at least one outbound transition.
\begin{align*}
	p(\text{stop}) &= \left\{
		\begin{array}{rl}
		\tau (1-\tau)^b & r_n \in F \wedge \exists a : \delta(r_n, a)\in Q
		\\ (1-\tau)^b & r_n \in F \wedge \not \exists a : \delta(r_n, a)\in Q
		\\ 0 & \text{otherwise}
		\end{array}
		\right.
\end{align*}

@@

Our equation for $p(\text{stop})$ is thus:
\begin{align*}
	p(\text{stop}) &= \left\{
		\begin{array}{rl}
		\frac{1}{b^2+3b+1} & r_n \in F \wedge \exists a : \delta(r_n, a)
		\\ \frac{1}{b+1} & r_n \in F \wedge \not \exists a : \delta(r_n, a)
		\\ 0 & \text{otherwise}
		\end{array}
		\right.
\end{align*}

\section{Experiment Design and Model Fitting}
Having generated the above Bayesian model for selection of the regular expression with the highest posterior probability, we then conducted a behavioral experiment.  41 subjects (Median age 34, std-dev 13.3) were surveyed on Amazon Mechanical Turk. Two groups (one of 20, one of 21) were presented with 10 yes or no questions in which they were asked to determine whether a given string `fit' with a list of strings. A sample question was as follows:
\begin{center}Given the list:

@@

\subsection{Parameter Fitting}
Because the role of $\beta$ had the effect of biasing models with fewer transitions, it also had a tendency to bias regular expressions with fewer states. For instance, the $\beta$ term would prefer the regular expression `a*' over `aaaa'. Thus, this term was fitted before fitting the $\alpha$ term.

@@

\begin{figure}[ht]
\begin{center}
\begin{align*}
\includegraphics[scale=0.5]{results.png}
\end{align*}
\end{center}
\caption{Chart showing  $p(s \in \mathcal{L} | S \subseteq {L}) $. Human subject data is in red, model data is in blue. $R$ = .75} 
\label{results-chart}
\end{figure}

>>>>

\subsection{Likelihood}
To define the likelihood of producing certain strings, we turn a DFA into a generative automaton with a probability distribution over generated strings. Letting the input strings be independent, the probability of generating the set of strings $S$ is equal to the product of the probabilities of generating each string.
\begin{align*}
	p(S|h) = \prod_{s \in S} p(s|h)
\end{align*}

Let the string $s$ be the concatenation of characters $\{a_0,a_1,...,a_{n-1}\}$. We define the probability of generating a string as the product of the probability of generating the letters in sequence $p(a_0,a_1,...,a_{n-1})$ and the product of stopping the string generation on an accept state $p(\text{stop}|h)$.
\begin{align*}
	p(s|h) = p(a_0,a_1,...,a_{n-1}|h)p(\text{stop}|h)
\end{align*}

We define the probability of generating the characters as the product of the probabilities of making each state transition.
\begin{align*}
	p(a_0, a_1, ..., a_{n-1}|h) = \prod_{i=0}^{n-1}p(\delta_h(r_i, a_i))
\end{align*}
where $\delta_h$ is the transition function of $h$, $r_i$ is the state reached after the first $i$ characters, and $r_0$ is the start state.

To simplify inference, we assume uniform probability over all possible outbound transitions for a given state. Formally, the probability of making a state transition is given by:
\begin{align*}
	p(\delta_h(r_i, a_i)) &= \frac{1}{\sum_{a\in A} 1\{\delta_h(r_i, a)\in Q_h\}}
\end{align*}
where $Q_h$ is the set of states in hypothesis $h$ and $1\{\cdot\}$ is the indicator function, which evaluates to $1$ if the enclosed proposition is true and $0$ otherwise. $\delta_h(r_i, a)\in Q_h$ indicates that the state transition leads to another state of the DFA, so the denominator counts the number of characters for which a transition is defined. In the case of the wildcard `A', the transition is defined for all 26 letters, so the denominator is 26.

@@

Instead of leaving $\tau$ as a parameter of the model, we use a technique explained by \citeA{HwangStuhlmullerGoodman2011}: assume a uniform prior over the parameter and marginalize it when calculating the likelihood of a string. So, let $b$ be the number of times the DFA skipped an accept state before accepting. That is, $b$ is given by the equation below, where $F_h$ is the set of accept states for hypothesis $h$.
\begin{align*}
	b = \sum_{i=0}^{n-1} 1 \{ r_i \in F_h\}
\end{align*}

The probability of stopping at an accept state is given by the equation below. The condition $r_n \in F_h \wedge \exists a : \delta_h(r_n, a)\in Q_h$ states that the last state reached by the string is an accept state, and that that state has at least one outbound transition.
\begin{align*}
	p(\text{stop}|h) &= \left\{
		\begin{array}{rl}
		\tau (1-\tau)^b & r_n \in F_h \wedge \exists a : \delta_h(r_n, a)\in Q_h
		\\ (1-\tau)^b & r_n \in F_h \wedge \not \exists a : \delta_h(r_n, a)\in Q_h
		\\ 0 & \text{otherwise}
		\end{array}
		\right.
\end{align*}

@@

Our equation for $p(\text{stop}|h)$ is thus:
\begin{align*}
	p(\text{stop}|h) &= \left\{
		\begin{array}{rl}
		\frac{1}{b^2+3b+1} & r_n \in F \wedge \exists a : \delta(r_n, a)
		\\ \frac{1}{b+1} & r_n \in F \wedge \not \exists a : \delta(r_n, a)
		\\ 0 & \text{otherwise}
		\end{array}
		\right.
\end{align*}

\section{Experiment Design and Model Fitting}
Having generated the above Bayesian model for selection of the regular expression with the highest posterior probability, we then conducted a behavioral experiment.  41 subjects (Median age 34, std-dev 13.3) were surveyed on Amazon Mechanical Turk. Two groups (one of 20, one of 21) were presented with 10 yes or no questions in which they were asked to determine whether a given string ``fit'' with a list of strings. A sample question was as follows:
\begin{center}Given the list:

@@

\subsection{Parameter Fitting}
Because the role of $\beta$ had the effect of biasing models with fewer transitions, it also had a tendency to bias regular expressions with fewer states. For instance, the $\beta$ term would prefer the regular expression \verb!a*! over \verb!aaaa!. Thus, this term was fitted before fitting the $\alpha$ term.

@@

\begin{figure}[ht]
\begin{center}
\begin{align*}
\includegraphics[scale=0.5]{results.png}
\end{align*}
\end{center}
\caption{Chart showing  $p(s \in \mathcal{L} | S \subseteq {L}) $, sorted in increasing order for model data. Human subject data is in blue, model data is in red. $R$ = .75} 
\label{results-chart}
\end{figure}
